$Z_3$ topological order in the quantum dimer-pentamer model

### 2016-08-08 qdpm_paper.tex
Worked on the appendix

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[tpb]
    \centering
    \includegraphics[width=1.0\columnwidth]{mag_loops_on_orntd_lat.pdf}
    \caption{$\Phi_i$: a nontrivial string winding around the system in one of the two possible
        directions. $\Phi_2$: an open
    magnetic string with two different charged (colored) visons at the end points. $\Phi_3$ a trivial closed
    string. The colors red and blue distinguish the two different sublattices that orient the links.}
    \label{fig:orientedSL}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

    \section{$Z_3$ Toric Code on the Square Lattice}
        In the $Z_3$ toric three degrees of freedom live on each link.
        We can represent the $Z_3$ algebra through the following expressions
        %
        %Z_N
        %\begin{equation}
        %    E |n\rangle = n |n\rangle, ~ n=0,1,2,...,N-1
        %    \\
        %    e^{iA} |n\rangle = |n+1 \rangle, ~ n=0,1,2,...,N-2
        %    \\
        %    e^{iA} |N-1\rangle = |0\rangle
        %    %\label{eqn:}
        %\end{equation}
        %Z_3
        %
        \begin{equation}
            \begin{split}
            & E_l |n_l\rangle = n_l |n_l\rangle, ~ n_l=0,1,2
            \\
            & e^{iA_l} |n_l \rangle = |n_l+1 \rangle, ~ n_l =0,1
            \\ 
            & e^{iA_l} |2 \rangle = |0\rangle
            \end{split}
            ,
            %\label{eqn:}
        \end{equation}
        %
        where the operators $E_l$ and $A_l$ act on a given link $l$ and are analogous to the 
        electric field and vector potential respectively. 
        By assigning orientations to the links between two vertices we can define the three degrees of
        freedom pictorially. 
        %
        \begin{itemize}
            \item $\vvemptylink$ 
            \item $\rightarrowlink$ 
            \item $\leftarrowlink$ 
        \end{itemize}
        % 
        To simplify notation, we define $e^{iA_l}$ as the operator $Q$ which acts on the links in the
        following way
        \begin{equation}
            \begin{split}
                & Q^\dagger_{vv'}  | \vvemptylink \rangle = | \rightarrowlink \rangle
                ,
                \\
                & Q^\dagger_{vv'}  | \rightarrowlink \rangle = | \leftarrowlink \rangle
                ,
                \\
                & Q^\dagger_{vv'}  | \leftarrowlink \rangle = | \emptylink \rangle
                .
            \end{split}
            %\label{eqn:}
        \end{equation}
        The link is specified by the two vertices it connects. The operator
        $Q^\dagger_{vv'}$ acts in the direction from the first subscript to the second subscript, in
        this case denoted as $v$ and $v'$. 
        Changing the direction in which $Q$ operates is equivalent to taking the hermitian conjugate
        of the operator withougt changing the direction.  
        %
        \begin{equation}
            (Q^\dagger_{vv'})^\dagger = Q_{vv'}^\dagger
            .
            %\label{eqn:}
        \end{equation}

        We now define an operator $E_{vv'}$ in a similar way, 
        %
        \begin{equation}
            \begin{split}
                & E^\dagger_{vv'}  | \vvemptylink \rangle = 0
                ,
                \\
                & E^\dagger_{vv'}  | \rightarrowlink \rangle = | \rightarrowlink \rangle
                ,
                \\
                & E^\dagger_{vv'}  | \leftarrowlink \rangle = 2| \leftarrowlink \rangle
                ,
            \end{split}
            %\label{eqn:}
        \end{equation}
        and with it define the last important operator that acts on the links. 
        %
        \begin{equation}
            P_{vv'}^\dagger \equiv e^{i2\pi E_{vv'}/3}
            .
            %\label{eqn:}
        \end{equation}
        %
        $P_{vv'}^\dagger$ acts on the links in the following way: 
        %
        \begin{equation}
            \begin{split}
                & P^\dagger_{vv'}  | \vvemptylink \rangle = | \vvemptylink \rangle
                ,
                \\
                & P^\dagger_{vv'}  | \rightarrowlink \rangle = e^{i2\pi/3} | \rightarrowlink \rangle
                ,
                \\
                & P^\dagger_{vv'}  | \leftarrowlink \rangle = e^{-i2\pi/3} | \leftarrowlink \rangle
                .
            \end{split}
            %\label{eqn:}
        \end{equation}
        %

        With the above definitions the $Z_3$ toric code Hamiltonian can be written in terms of the $P$ and $Q$ operators.
        In general a string net Hamiltonian might be written as 
        \begin{equation}
            H = - J_e \sum_v A_v - J_m \sum_p B_p
            %\label{eqn:}
        \end{equation}
        where 
        \begin{equation}
            A_v = \prod_{i\in v} P^\dagger_{vv_i}~,~ B_p = \prod_{i\in p} Q^\dagger_{v_iv_{i+1}}
            .
            %\label{eqn:}
        \end{equation}
        The specific vertex and plaquette labeling scheme is shown in Fig.~\ref{fig:vertex_link_labels}.
        %
        \begin{figure}[htpb]
            \centering
            \includegraphics[width=0.8\linewidth]{vertex_link_gauge_def}
            \caption{Notation of vertex and plaquette labels}
            \label{fig:vertex_link_labels}
        \end{figure}
        %
        For the ground state of this Hamiltonian there is aa local symmetry corresponding to the
        minimization of the first term. This local symmetry is a gauge symmetry that can be
        expressed in terms of the gauge operator.
        \begin{equation}
            G_v | \psi \rangle= | \psi \rangle
            %\label{eqn:}
        \end{equation}
        where $G_v$ = $A_v$.

        \subsubsection{$Z_3$ String operators}
            With the above notation we can now write down the $Z_3$ string operators. The electric string operator along
            the path $\Gamma$ is written
            %
            \begin{equation}
                S^e_{\Gamma} = \prod_{l=0}^{N-2} Q_{v_lv_{l+1}}^\dagger
                %\label{eqn:}
                ,
            \end{equation}
            %
            where $v_l \in \Gamma$ and $N$ is the number of vertices in $\Gamma$. We show an example of an
            electric string operator operatiing on an example configuration in Fig.~\ref{fig:example_elec_string}.
            %
            %
            \begin{figure}[htpb]
                \centering
                \includegraphics[width=0.8\linewidth]{example_elec_string.pdf}
                \caption{The grey line is the path $\Gamma$ on which the string operator acts. The operator acts
            in the direction from the vertex labeled in red to the one labeled in blue.}
                \label{fig:example_elec_string}
            \end{figure}
            %
            %
            The magnetic string operator can be written as
            \begin{equation}
                S^m_{\Gamma} = \prod_{l=0}^{N-1} P_{v_{2l}v_{2l+1}}^\dagger
                %\label{eqn:}
            \end{equation}
            %
            where $v_l$ are the vertices adjacent to the links in $\Gamma$ and $N$ is the number of
            vertices
            on the \textit{left} side of the path when looking along the path from red to blue. 
            An example of this operator acting
            on a configuration is shown in Fig.~\ref{fig:example_mag_string} along with the vertex labeling
            scheme.
            %
            %
            \begin{figure}[htpb]
                \centering
                \includegraphics[width=0.8\linewidth]{example_mag_string.pdf}
                \caption{The grey line is the path $\Gamma$ on which the string operator acts. The magnetic
                    string operator acts in the 
                    direction from the plaquette labeled in red to the one labeled in blue.}
                \label{fig:example_mag_string}
            \end{figure}
            %
            The magnetic string operator acting on path $\Gamma$ in Fig.~\ref{fig:example_mag_string} is
            %
            \begin{multline}
                S^m_{\Gamma} (\mathrm{~Fig.~\ref{fig:example_mag_string}~config})
                = P_{v_0v_1}^\dagger P_{v_2v_3}^\dagger P_{v_4v_5}^\dagger
                \\
                = \exp{[0 - \frac{i2\pi}{3} + \frac{i2\pi}{3} ]} 
                = 1 (\mathrm{~Fig.~\ref{fig:example_mag_string}~config})
                .
            \end{multline}
            %

            An interesting property of the electric string operator in the $Z_3$ toric code is
            that an unclosed string produces two different defects. 
            An example of an open string and the defects is shown in Fig.~\ref{fig:example_elec_string}. The
            configuration in Fig.~\ref{fig:example_elec_string}(a) obeys the local gauge symmetry
            aside from the endpoints of the path that it acts on ($\Gamma$). The two defects at the
            end points produce two different values when the vertecies are operated on by $G_v$, being
            $G_{\mathrm{red}} = e^{i2\pi /3}$, and $G_{\mathrm{blue}} = e^{-i2\pi /3}$.

            The magnetic excitations, or $Z_3$ visons live at the endpoints of the magnetic
            strings.  These visons can change the phase of a configuration by different values.

%        \subsubsection{properties of $Z_3$ string operators}

        \subsubsection{Winding Numbers}

            The ground state of the $Z_3$ toric code has nine-fold topological degeneracy because of
            the $3\times3$ different topological sectors. As the case of the toric code the $Z_3$
            electric string changes the topological sector when acting on on a nontrivial closed
            path. We call the electric winding
            operator that acts around the major axis $W^e_y$ and around the minor axis $W^e_x$.
            The topological sector can be determined 
            by acting around the axes with the magnetic string operator, called $W^m_y$ and
            $W^m_x$. All of the winding commute with the Hamiltonian and
            therefore share the same eigenvalues. The operator $W^m_{x}$ ($W^m_{y}$) can 
            have three different values depending on how many electric strings enclose the minor
            (major) axis. The three different
            possible values for each winding number around each axis gives the model nine distinct
            topological sectors.

\section{Additional numerical results on the QDPM RK ground state}

%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}
    \centering
    \includegraphics[width=1.0\columnwidth]{spatial_cors_end_to_end_parallel.pdf}
    \caption{Correlation function between parallel dimers (end-to-end) on the same (blue) and different (red) 
    sub-lattice for a $L=64$ lattice.}
    \label{fig:spatial_dmr_cor}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[tpb]
    \centering
    \includegraphics[width=1.0\columnwidth]{mag_loops_on_orntd_lat.pdf}
    \caption{$\Phi_i$: a nontrivial string winding around the system in one of the two possible
        directions. $\Phi_2$: an open
    magnetic string with two different charged (colored) visons at the end points. $\Phi_3$ a trivial closed
    string. The colors red and blue distinguish the two different sublattices that orient the links. \note{Owen: Add a dimer/pentamer-ization to the figure. Also, add a monomer defect in the magnetic loop, to demonstrate the fractional statistics of defects.}}
    \label{fig:orientedSL}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{$Z_3$ Gauge theory on the square lattice}

Here we will review a string-net implementation of $Z_3$ gauge theory on the square lattice~\cite{Horn1979}. We define a three -dimensional Hilbert space on each link, corresponding to the presence or absence of an oriented string segment that carries $Z_3$ electric flux which are eigenstates of the Hermitian $Z_3$ electric flux operator $E_{vv'}$
\begin{align}
        & E_{vv'}  \eket{\vvemptylink} = 0
        ,
        \\
        & E_{vv'}  \eket{\rightarrowlink} = + \eket{\rightarrowlink}
        ,
        \\
        & E_{vv'} \eket{\leftarrowlink} = - \eket{\leftarrowlink}
        .
\end{align}
\note{Owen:can you add symbols with $v$ and $v'$ at both end points for all types of links?}.

At each vertex we define a local unitary gauge transformation $G_v$,
\begin{equation}
G_v \equiv \prod_{v'} P^\dagger_{vv'}
\end{equation}
where the product is over all nearest neighbor vertices $v'$, and we have defined the unitary operator
\begin{equation}
P_{vv'}^\dagger \equiv e^{i 2\pi/3 E_{vv'}}.
\end{equation}
We consider eigenstates of $G_v$ to have a definite $Z_3$ electric charge at vertex $v$. The charge-free sector is the gauge invariant subspace where
\begin{equation}
G_v \eket{\psi} = \eket{\psi} \quad \forall v.
\end{equation}
This gauge invariant subspace corresponds to states where the ``string flux" vanishes modulo 3 at each vertex, thus including vertices where three inward or three outward strings meet at a vertex. For eigenstates of $G_v$ with  nonzero electric charge
\begin{equation}
G_v \ket{\psi} = e^{\pm 2\pi/3 i} \ket{\psi}
\end{equation}
which corresponds to the presence of a $Z_3$ charge of $\pm1$ at vertex $v$. A vertex with a positive (negative) charge has a net outward (inward) string flux of 1 modulo 3, or net inward (outward) flux of 2. \figref{fig:example_elec_string} shows examples of charged and charge free string configurations.

To provide dynamics, we define a unitary operator electric flux ladder operator $Q^\dagger_{vv'}$ that increments the electric flux:
\begin{align}
        & Q^\dagger_{vv'}  \eket{\vvemptylink} = \eket{\rightarrowlink}
        ,
        \\
        & Q^\dagger_{vv'}  \eket{\rightarrowlink} = \eket{\leftarrowlink}
        ,
        \\
        & Q^\dagger_{vv'}  \eket{\leftarrowlink} = \eket{\emptylink}
\end{align}
The simplest gauge invariant Hamiltonian is then
\begin{equation}
H_{Z_3} = - J  \sum_p  \left( Q^\dagger_{v_1^pv_2^p} Q^\dagger_{v_2^pv_3^p} Q^\dagger_{v_3^pv_4^p} Q^\dagger_{v_4^pv_1^p} + h.c. \right)
\end{equation}
where $v_i^p$ represent the four vertices in the plaquette $p$, ordered cyclically by nearest neighbors.

We can define two oriented open string operators which commute with $H_{Z_3}$ everywhere except at the endpoints. The electric string operator $S_e$, which acts on an open, oriented string on the lattice $\gamma_{vv'}$ with endpoints $v,v'$, and creates a $+1$ ($-1$) charge at vertex $v$ ($v')$, is defined as
\begin{equation}
    S_e \left( \gamma_{vv'} \right) \equiv \prod_{l \in \gamma_{vv'}} Q_{v_l v_l'}^\dagger
\end{equation}
where the product is over all links $l$ in $\gamma_{vv'}$, $v_l$ and $v_l'$ are the two vertices in $l$, ordered according to the orientation of $\gamma_{vv'}$ (see \figref{fig:example_elec_string}). The magnetic string operator $S_m$ is defined by an open, oriented string on the dual lattice and creates a magnetic vortex-antivortex pair on plaquettes at the endpoints of $\tilde{\gamma}_{pp'}$
\begin{equation}
    S_m \left( \tilde{\gamma}_{vv'} \right) \equiv \prod_{l \perp \tilde{\gamma}_{pp'}} P_{v_l v_l'}^\dagger
\end{equation}
where the product is over all links $l$ that cross $\tilde{\gamma}$ and $v_l$ ($v_l'$) is the vertex to the left (right) of the string as $\tilde{\gamma}$ is traversed from $p$ to $p'$. 
%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[t]
    \centering
    \includegraphics[width=0.8\linewidth]{example_elec_string.pdf}
    \caption{ ({\it left}) An example of a charge free $Z_3$ electric string configuration. The electric string operator $S_\gamma^e$ represented by the thick gray line, oriented from the red to blue vertex. ({\it right}) An example of a $Z_3$ electric string configuration with a $+1$ ($-1$) electric charge at the red (blue) vertex; this string configuration results form the action of $S_\gamma^e$ on the left panel. \note{Owen: Can you remove the a) and b) labels on the figure? Also switch the $\Gamma$ to $\gamma$, and remove the $v_0-v_3$ labels. Add a gray arrow on the string operator from red to blue. Can you add a non-trivial magnetic loop around the blue charge on the right?} }
    \label{fig:example_elec_string}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[]
    \centering
    \includegraphics[width=0.5\columnwidth]{example_mag_string.pdf}
    \caption{The magnetic string operator $S_m$ is defined by the open, oriented string $\tilde{\gamma}_{pp'}$, and creates a vortex (anti-vortex) a the red (blue) plaquette. \note{Owen: can you add a gray arrow orienting the string from red to blue. Change the label to $\tilde{\gamma}$. Maybe we should distinguish between the electric and magnetic charges in these two figures (maybe add a small e/m in the circles?}}
    \label{fig:example_mag_string}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%

Note that  $S_e$ and $S_m$ do not commute if their corresponding strings intersect--this underlies the non-trivial statics of electric charges and magnetic vortices. Consider the magnetic string operator $S_m(\tilde{\Gamma})$ defined by an oriented \emph{closed} loop $\tilde{\Gamma}$ on the dual lattice. This operator corresponds to the creation of a vortex-antivortex pair, winding of the vortex around $\Gamma$, and annihilation of the vortex pair. A state with definite electric charge inside $\Gamma$ will be an eigenstate of $S_m(\tilde{\Gamma})$, with eigenvalue $e^{ i 2 \pi/3 q}$, where $q=0,\pm1$ is the net $Z_3$ electric charge (modulo 3) inside $\tilde{\Gamma}$. For $\Gamma$ winding around an isolated electric charge, we see that electric charges and magnetic vortices are anyons with relative braiding statics with a statistical phase of $e^{\pm i 2\pi/3}$.

Additionally, topologically non-trivial string operators give rise to a topological degeneracy...

%\section{Additional numerical results on the QDPM RK ground state}
%
%%%%%%%%%%%%%%%%%%%%%%%%
%\begin{figure}
%    \centering
%    \includegraphics[width=1.0\columnwidth]{spatial_cors_end_to_end_parallel.pdf}
%    \caption{Correlation function between parallel dimers (end-to-end) on the same (blue) and different (red) 
%    sub-lattice for a $L=64$ lattice.}
%    \label{fig:spatial_dmr_cor}
%\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%
